A Framework for Equivalence of Cast Calculi and a new Cast Representation

### 2020-10-12 paper.tex
Updates from Overleaf
a general description of the proof of lemma 3

\begin{document}
%
\title{A Framework for Equivalence of Cast Calculi and a new Cast Representation}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Kuang-Chen Lu\inst{1} \and
Jeremy G. Siek\inst{2} \and
Andre Kuhlenschmidt\inst{2}
}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Brown University, Providence RI 02912, USA \and
Indiana University, Bloomington IN 47405, USA}
% \institute{Princeton University, Princeton NJ 08544, USA \and
% Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
% \email{lncs@springer.com}\\
% \url{http://www.springer.com/gp/computer-science/lncs} \and
% ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
% \email{\{abc,lncs\}@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Gradual typing combines dynamic typing and static typing in one
language to provide the advantages of both typing
disciplines. Implementing gradual typing with reasonable overhead,
however, is not trivial. Casts accumulate on higher-order values and
continuations, changing the overall space complexity. Several
space-efficient cast representations have been proposed, notably
coercions in normal forms and threesomes. But they are presented in
different dynamic semantics, which make it difficult to compare the
key differences between the cast representations. We present a
abstract data type, namely Cast ADT, that captures interface functions
a cast representations typically includes, and an abstract machine
parameterized over Cast ADTs.  The Cast ADT brings another benefit:
space-efficient cast representations shall be proved not changing the
overall behavior of language implementations except saving space.
With our generic abstract machine, these proofs can be done by proving
properties of a specific Cast ADT, which is simpler than proofs
regarding the entire dynamic semantics.
\todo[inline]{JS: The stuff above about ``make it difficult to compare''
  is a weak argument. I would recommend deleting that argument and
  instead focus on the benefits regarding the proofs of correctness.}

We have successfully applied our new approach to the proofs of
correctness for two space-efficient cast representations from the
literature
(\cite{Siek:2010:TWB:1706299.1706342,siek2012interpretations,Siek:2015:BCT:2737924.2737968})
and believe this framework is also applicable to other cast
representations
(\cite{Garcia:2013:CTB:2500365.2500603,siek2012interpretations}).  We
also present a new cast representation, hypercoercions, that is more
friendly to mechanized meta-theory and consumes less memory than
coercions in normal forms. We prove correctness for two variants of
hypercoercions.

\todo[inline]{KC: I feel ``going through a bisimulation proof'' is too specific.
There should be other ways to prove equivalence of two dynamic semantics.
How does Jeremy think?}
\todo[inline]{JS: yes}

@@

\todo[inline]{KC: Cite "Contracts for higher-order functions" for the proxy-based approach maybe?}
\todo[inline]{JS: That paper doesn't exactly fit into what you're talking about in the above
sentence. It might be appopriate to cite it elsewhere in the paper. Also,
there are earlier papers about coercive subtyping (Computing with coercions by Breazu-Tannen et al)
that use a proxy-based approach. The contracts paper may be the first to propose blame tracking.}
\todo[inline]{KC: How about citing the origin of the proxy-based approach when we talk
about the CPSed even-odd example?
I am confused about which paper to cite. \cite{herman2010space} seems to be citing
\cite{findler2002contracts} for the proxy-based approach of casting function.
Here is the relevant sentences from \cite{herman2010space}:
``In general, it is not possible to check if an untyped function satisfies a
particular static type. A natural solution is to wrap the function in a proxy that, whenever
it is applied, casts its argument and result values appropriately, ensuring that the function
is only observed with its expected type. This proxy-based approach was formalized by Findler
and Felleisen [15] ...''
I am \textit{not} trying to say I don't believe in your statements about the origins of
the proxy-based approach and blame tracking in functions.
I want to know whether \citet{herman2010space} is citing the right paper.
Maybe both papers came up independently with a similar idea but \cite{findler2002contracts}
is more related to our work.
}
\todo[inline]{JS: In the above you've lumped proxy-based and blame
  tracking together, which is incorrect. AFAIK,
  \cite{findler2002contracts} is the origin of blame tracking. AFAIK,
  \citet{Breazu-Tannen:1990vc} is the origin of proxy-based casts.
  You can read both papers for yourself and be the judge.  BTW, TAPL
  (section 15.6) cites Breazu-Tannen.  So \citet{herman2010space}
  cited the wrong paper for function proxies, perhaps because they
  didn't know about \citet{Breazu-Tannen:1990vc}. Also,
  \cite{findler2002contracts} should have cited
  \citet{Breazu-Tannen:1990vc} but did not. Perhaps because they did
  not consider that papers about ``casts'' to be closely related to
  their paper about ``contracts''.  Or perhaps they too did not know
  about \citet{Breazu-Tannen:1990vc}.  Authors make mistakes all the
  time regarding related work, and this is just yet another example of
  that. For that matter, I should have cited
  \citet{Breazu-Tannen:1990vc} in \citet{siek2006gradual} but did not,
  probably because I had learned about it from TAPL and therefore
  thought it was so standard that it didn't need a citation.  But it
  would have been better for me to cite it.}

@@

\todo[inline]{KC: in the next paragraph I cite \cite{breazu1991inheritance}
and \cite{breazu1990computing}}

Continuations are not the only place where casts can accumulate.
Consider a variant of the above program written in
continuation-passing style.  Here, both functions returns
\lstinline|Bool| so there is no implicit cast in tail position.
% However, the argument \lstinline|k| is wrapped in a cast in each
% recursive call because the type of \lstinline|k| is less precise in
% \lstinline|even?| than it is in \lstinline|odd?|.
However, the argument \lstinline|k| is casted in each
recursive call because the type of \lstinline|k| is less precise in
\lstinline|even?| than it is in \lstinline|odd?|. 
Every time a function is casted we record a pieces of
cast information for the input and another piece for
the output, following the approach in \cite{breazu1991inheritance}
and \cite{breazu1990computing},
so the space complexity is again linear.
\begin{lstlisting}
(define (even? [n : Int] [k : (-> $\star$ Bool)]) : Bool
  (if (= n 0) (k #t) (odd? (- n 1) k)))
(define (odd? [n : Int] [k : (-> Bool Bool)]) : Bool
  (if (= n 0) (k #f) (even? (- n 1) k)))
\end{lstlisting}
These example programs are adapted from those of \cite{herman2010space}.
We direct readers to their work for more examples and a detailed explanation.

@@

\todo[inline]{KC: how can I forget the interpretation paper!}
\cite{siek2012interpretations} presents a definitional interpreter
and a variant of the ECD abstract machine \citep{siek2012ECD}.
Both systems are parameterized over 4 flavors of coercions in normal
forms: Lazy UD, Lazy D, Eager UD, Eager D.

@@

\todo[inline]{JS: The above is too short of a description of that related
work.}
\todo[inline]{KC: I am actually very confused by \citet{garcia2014deriving}.
They stated that ``Theorems 8.1 and 8.3 can be generalised for other choices
of dynamic semantics by applying the correspondence described
through the paper starting with a different set of coercion artefacts.''.
However they didn't give a generalization and it is unclear to me over what
their system is parameterized (cast representations? apply-coercion and seq?.}
\todo[inline]{JS: your comment above says what you should say
  (except the part about your confusion). Say it in a matter-of-fact
way, and not a derogitory way.}
\todo[inline]{KC: updated.}

@@

We come to the main lemma, the weak bisimulation between
\CBMachine\ and \SMachine{C}.

\begin{lemma}[Weak Bisimulation between \CMachine{B} and \SMachine{C}] \label{lem:weak-bisimulation}
Assume
\begin{enumerate}
    \item the blame strategy $B$ and $C$ agrees on the set of injectable types and
    \item $\denote{c}(v)$ preserves the bisimulation relation (e.g. Lemma~\ref{lem:LazyD-apply-cast-preserve-bisim} and \ref{lem:LazyUD-apply-cast-preserve-bisim}) and
    \item $s_1$ is a state in \CMachine{B} and
    \item $s_1'$ is a state in \SMachine{C} and
    \item $\bisims{s_1}{s_1'}{T}$
\end{enumerate}	then either
\begin{enumerate}
\item $s_1  \text{ final}$ and $s_1' \text{ final}$ or
\item $\reduceto{s_1}{^{+}s_2}$
  and $\reduceto{s_1'}{^{+}s_2'}$
  and $\bisims{s_2}{s_2'}{T}$
\end{enumerate}
where the final predicate on states is defined as follows
\begin{gather*}
\infr{}{(\error{l})\text{ final}} \quad
\infr{}{(\shalt{v})\text{ final}}
\end{gather*}
\end{lemma}
\todo[inline]{KC: wouldn't it be great if I can describe the main ideas of the proof?}
\todo[inline]{JS: Yes, that would be great.}
\todo[inline]{KC: that is possibly only if we steal some pages from elsewhere.
The paper limit is 25 (excluding bib) and we have used 30 now.
https://etaps.org/2021/call-for-papers}
\todo[inline]{KC: [IMPORTANT]}

@@

\todo[inline]{JS: I'd recommend moving this subsection before that of
  LazyD because a reader is more likely to be familiar with Lazy UD.}
\todo[inline]{KC: Updated.}

@@

\todo[inline]{KC: I need a better name for \compatibletailhead{t}{h},
  any suggestion from Jeremy?}
\todo[inline]{JS: how about $\sim$}
\todo[inline]{KC: sounds good}

@@

\todo[inline]{KC: should we cite the place where ``Prior presentations'' locate?}
The translation function \translate{\ccast{S}{l}{T}} is defined in 
\figref{fig:LazyUD-Hypercoercions-translation}.  Prior
presentations of this function do not use auxiliary functions, as we
do here. The reason that we introduce the auxiliary functions
$\mathrm{castToDyn}$ and $\mathrm{castFromDyn}$, is to ensure that each of them 
is structurally recursive, which makes them straightforward to define in
Agda.

@@

>>>>

\begin{document}
%
\title{A Framework for Equivalence of Cast Calculi and a new Cast Representation}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Kuang-Chen Lu\inst{1} \and
Jeremy G. Siek\inst{2} \and
Andre Kuhlenschmidt\inst{2}
}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Brown University, Providence RI 02912, USA \and
Indiana University, Bloomington IN 47405, USA}
% \institute{Princeton University, Princeton NJ 08544, USA \and
% Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
% \email{lncs@springer.com}\\
% \url{http://www.springer.com/gp/computer-science/lncs} \and
% ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
% \email{\{abc,lncs\}@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
% Background
Gradual typing combines dynamic typing and static typing in one
language to provide the advantages of both typing
disciplines. Implementing gradual typing with reasonable overhead,
however, is not trivial. Casts accumulate on higher-order values and
continuations, changing the overall space complexity. Several
space-efficient cast representations have been proposed, notably
coercions in normal forms and threesomes.
These cast representations shall be proved not changing the overall
behavior of language implementations except saving space.
Existing proofs typically go through a bisimulation between
a space-efficient semantics, which use a specific cast representation,
and the definitional but inefficient semantics \citep{Siek:2010:TWB:1706299.1706342,Siek:2015:BCT:2737924.2737968}.
% Problem
However, this technique imposes unnecessary burden to proof writer
because a considerable portion of the semantics is irrelevant to
casts.
% Solution
To isolate cast representations from the remaining part of 
implementations, we present a abstract data type, namely Cast ADT,
that captures interface functions a cast representations typically 
includes, and an abstract machine parameterized over Cast ADTs.
Our generic dynamic semantics comes with generic theorems,
% Why is our solution a step forward
with which the correctness proofs can be done by just proving 
properties of a specific Cast ADT.
% %
% To isolate cast representations from the remaining part of 
% implementations, we present a abstract data type, namely Cast ADT,
% that captures interface functions a cast representations typically 
% includes, and an abstract machine parameterized over Cast ADTs.
% %
% With our generic abstract machine, the correctness proofs can be
% done by proving properties of a specific Cast ADT, which is simpler
% than proofs regarding the entire dynamic semantics.
%%%%%
% %
% But they are presented in
% different dynamic semantics, which make it difficult to compare the
% key differences between the cast representations.
% %
% We present a
% abstract data type, namely Cast ADT, that captures interface functions
% a cast representations typically includes, and an abstract machine
% parameterized over Cast ADTs.
% %
% The Cast ADT brings another benefit:
% space-efficient cast representations shall be proved not changing the
% overall behavior of language implementations except saving space.
% %
% With our generic abstract machine, these proofs can be done by proving
% properties of a specific Cast ADT, which is simpler than proofs
% regarding the entire dynamic semantics.
\todo[inline]{JS: The stuff above about ``make it difficult to compare''
  is a weak argument. I would recommend deleting that argument and
  instead focus on the benefits regarding the proofs of correctness.}
\todo[inline]{KC: updated.}

We have successfully applied our new approach to the proofs of
correctness for two space-efficient cast representations from the
literature
\citep{Siek:2010:TWB:1706299.1706342,siek2012interpretations,Siek:2015:BCT:2737924.2737968}
and believe this framework is also applicable to other cast
representations
\citep{Garcia:2013:CTB:2500365.2500603,siek2012interpretations}.  We
also present a new cast representation, hypercoercions, that is more
friendly to mechanized meta-theory and consumes less memory than
coercions in normal forms. We prove correctness for two variants of
hypercoercions.

@@

@@

\todo[inline]{KC: in the next paragraph I cite \cite{breazu1991inheritance}
and \cite{findler2002contracts}. My understanding is \cite{findler2002contracts}
extends the approach in \cite{breazu1991inheritance} with
blame labels.}

Continuations are not the only place where casts can accumulate.
Consider a variant of the above program written in
continuation-passing style.  Here, both functions returns
\lstinline|Bool| so there is no implicit cast in tail position.
% However, the argument \lstinline|k| is wrapped in a cast in each
% recursive call because the type of \lstinline|k| is less precise in
% \lstinline|even?| than it is in \lstinline|odd?|.
However, the argument \lstinline|k| is casted in each
recursive call because the type of \lstinline|k| is less precise in
\lstinline|even?| than it is in \lstinline|odd?|. 
Every time a function is casted we record a pieces of
cast information for the input and another piece for
the output, following the approach in \cite{breazu1991inheritance}
and \cite{findler2002contracts},
so the space complexity is again linear.
\begin{lstlisting}
(define (even? [n : Int] [k : (-> $\star$ Bool)]) : Bool
  (if (= n 0) (k #t) (odd? (- n 1) k)))
(define (odd? [n : Int] [k : (-> Bool Bool)]) : Bool
  (if (= n 0) (k #f) (even? (- n 1) k)))
\end{lstlisting}
These example programs are adapted from those of \cite{herman2010space}.
We direct readers to their work for more examples and a detailed explanation.

@@

\cite{siek2012interpretations} presents a definitional interpreter
and a variant of the ECD abstract machine \citep{siek2012ECD}.
Both systems are parameterized over 4 flavors of coercions in normal
forms: Lazy UD, Lazy D, Eager UD, Eager D.

@@

@@

We come to the main lemma, the weak bisimulation between
\CBMachine\ and \SMachine{C}.
A general description of the proof is in the Appendix.

\begin{lemma}[Weak Bisimulation between \CMachine{B} and \SMachine{C}] \label{lem:weak-bisimulation}
Assume
\begin{enumerate}
    \item the blame strategy $B$ and $C$ agrees on the set of injectable types and
    \item $\denote{c}(v)$ preserves the bisimulation relation (e.g. Lemma~\ref{lem:LazyD-apply-cast-preserve-bisim} and \ref{lem:LazyUD-apply-cast-preserve-bisim}) and
    \item $s_1$ is a state in \CMachine{B} and
    \item $s_1'$ is a state in \SMachine{C} and
    \item $\bisims{s_1}{s_1'}{T}$
\end{enumerate}	then either
\begin{enumerate}
\item $s_1  \text{ final}$ and $s_1' \text{ final}$ or
\item $\reduceto{s_1}{^{+}s_2}$
  and $\reduceto{s_1'}{^{+}s_2'}$
  and $\bisims{s_2}{s_2'}{T}$
\end{enumerate}
where the final predicate on states is defined as follows
\begin{gather*}
\infr{}{(\error{l})\text{ final}} \quad
\infr{}{(\shalt{v})\text{ final}}
\end{gather*}
\end{lemma}

@@

@@

@@

\todo[inline]{KC: should we cite the place where ``Prior presentations'' locate?}

The translation function \translate{\ccast{S}{l}{T}} is defined in 
\figref{fig:LazyUD-Hypercoercions-translation}.  Prior
presentations of this function do not use auxiliary functions, as we
do here. The reason that we introduce the auxiliary functions
$\mathrm{castToDyn}$ and $\mathrm{castFromDyn}$, is to ensure that each of them 
is structurally recursive, which makes them straightforward to define in
Agda.

@@

\todo[inline]{KC: The general idea of Lemma~\ref{lem:weak-bisimulation}}

Lemma~\ref{lem:weak-bisimulation} is proved by case analysis on \bisims{s_1}{s_1'}{T}.
Function call is the most interesting case. In this case the function value in
\CMachine{B}
must be of the form 
\[
v_1 = \vfuncc{
		\vfuncc{
			\vfunc{x}{S_1}{e}{E}
		}{S_1}{T_1}{l_1}{S_2}{T_2}
		\ldots
	}{S_{n-1}}{T_{n-1}}{l_{n-1}}{S_n}{T_n}
\]
By induction of the sequence of casts, we know \scont{v_2}{\Kcons{\FappOX{v_1}}{K}} will
transition to
\[
\begin{array}{l}
\mlet{v_2}{\applyCast{v_2}{\ccast{S_n}{\bar{l}}{S_{n-1}}}}{} \\
\ldots \\
\mlet{v_2}{\applyCast{v_2}{\ccast{S_2}{\bar{l}}{S_1}}}{} \\
\sexpr{e}{E[x:=v_2]}{
	\Kcons{\Fcast{\ccast{T_1}{l}{T_2}}}{
	    \ldots
	    \Kcons{\Fcast{\ccast{T_{n-1}}{l}{T_n}}}{K}}
}
\end{array}
\]
% Assuming \bisimv{v_1}{v_1'}{\Pfunc{S_n}{T_n}} and \bisimv{v_2}{v_2'}{S_n}.
As for \SMachine{C}, the function value must be of the form
\[
v_1' = \ecast{\vfunc{x}{S}{e}{E'}}{\Pfunc{c}{d}}
\]
where
\[
\begin{array}{l}
\bisimc{(\ccast{S_n}{\bar{l}}{S_{n-1}} \ldots \ccast{S_2}{\bar{l}}{S_1})}{c}{S_n}{S_1} \\
\bisimc{(\ccast{T_1}{l}{T_1} \ldots \ccast{T_{n-1}}{l}{T_n})}{d}{T_1}{T_n} \\
\end{array}
\]
By the definition of \applyContName\ we know \applyCont{v_2'}{\Kcons{\FappOX{v_1'}}{K}} will
transition to
\[
\begin{array}{l}
\mlet{v_2'}{\denote{c}(v_2')}{} \\
\sexpr{e}{E'[x:=v_2']}{\extCont{d}{K'}}
\end{array}
\]
Intuitively, we need to show the sequence of let bindings in \CMachine{B}
bisimulate the single let binding in \SMachine{C}, and that the two 
extended continuations bisimulate. Both of them are straightforward to
proof.
